Visualizing Conditional Dependencies

### 2018-01-31 paper/main.tex
newest version + repo clean up

@@

\usepackage{array,multirow,graphicx}
\usepackage{subfiles}
\usepackage{url}
\usepackage{amsmath}

\title{Visualizing Conditional Dependencies}
\author{Hannah Aizenman}
\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
\submitted{}
\abstract{Many datasets are too large or complex to discern patterns simply through visualizing the observations in them, so researchers instead compute the distributions of select variables and explore under what conditions they occur. While there are many visualization techniques that preserve variable type, dimensionality, and mutlivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships. This survey presents a sampling of techniques for visualizing distributions and densities conditioned on variables that are categorical or quantitative, discrete or continuous, and potentially have or infer dimensional dependencies. 
}
\begin{document}
\makefrontmatter

\section{Introduction}
\label{sec:introduction}
Many datasets are multivariate and often contain a a mix of quantative and categorical variables. The Iris dataset \cite{fisher_use_1936-1, _uci_????} is a well known example of the type of data commonly found in machine learning literature. It is so wildely used as a sample dataset because it has 150 observations that have both quantative and categorical attributes. The version visualized in the introduction to this paper is obtained from the scikit-learn Python machine learning library \cite{scikit-learn}. The Iris dataset contains the following attributes:\\
\begin{tabular} {lllll}
Sepal length (cm) & Sepal width (cm) & Petal length (cm) & Petal width (cm) & Species name
\end{tabular}
\\
A first step in exploring the dataset is typically to make a visualization, but of what sort? Bertin posits that all visualizations at their core have two parts \cite{bertin_semiology_2011}:
\begin{description}
\item[invariant] - invariable common ground
\item[components] - a finite set of variational concepts
\end{description}

@@

\begin{figure}
\includegraphics[width=1\textwidth]{intro/retinal_variables.png}
\caption{Bertin's retinal variables are a codification of how position, size, shape, color and texture can be used to illustrate variations in the components of a visualization. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} reproduced from Making Maps: A Visual Guide to Map Design for GIS \cite{krygier_making_2005}}
\label{fig:retinal_variables}
\end{figure}

@@

@@

\begin{figure}
\includegraphics[width=\textwidth]{intro/iris_parallel}
\caption{The parallel coordinates plot (PCP) of the Iris dataset shows the pairwise relationship between adjacent variables, for example sepal length and sepal width. The PCP plot can be used as proxy for illustrating conditional probability because it shows how the quantative measurements distribute amongst the species. This is is seen both in the relative widths of each band of color and in the similarites in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands.}
\label{fig:iris_parallel}
\end{figure}

While the Iris dataset does not have any variables that seem to traditionally fit into the model of keys-as it has no time or space variables-an analogous look-up task for the Iris dataset is to filter on a component and then look at the probabiluty distributions of the other variables in relation tp that key. In effect this treats all variables as potential keys into distributions of the other variables. In the parallel coordinates plot (PCP) in figure~\ref{fig:iris_parallel.png}, each line is an observation in the dataset, and each color maps to a species of Iris. The color coding then allows for this visualization to be used to show how the four quantative variables (sepal length, sepal width, petal length, and petal width) distribute within each variable.

@@

wherein A is the observation vector $\left< \text{sepal length, sepal width, petal length, petal width} \right>$ and B in turn takes on each species of petal. The locations of the color and width of the colored band act as proxies for the distribution of each quantative variable and the overall observation vector.

\begin{figure}
\includegraphics[width=\textwidth]{intro/iris_observations.png}
\caption{The diaganol of the scatter matrix shows the distribution of each variable relative to the species type. This can be used to illustrate the conditional probability of each variable relative to a type of species. The scatter matrix plot shows every set of pairwise relationships in the dataset, color code by Iris species. Because the scatter matrix shows the cooccurance of two variables and is encoded with information about a third variable, it works well was a tool to explore a potential conditional dependency amongst the variables.}
\label{fig:iris_observations}
\end{figure}

As exemplified by the scatter matrix in figure~\ref{fig:iris_observations}, there are a number of visualization techniques that can be used to explore conditional dependencies, but these methods are limited to datasets where the observations are discrete. There are visualization techniques for understanding the probability distributions of functional observations that could potentially be leverged in conjunction with these methods to develop methods for visualizing conditional dependencies in a functional space. This paper provides a sampling of the following visualizations:
\begin{enumerate}
\item methods for visualizing conditional dependency for discrete observations
\item methods for visualizing uncertainity and probabilistic visualizations for functional observations
\end{enumerate}

%%\subfile{sections/distdist.tex}
%%\subfile{sections/distdens.tex}
%%\subfile{sections/densdist.tex}
%%\subfile{sections/densdens.tex}

@@

1 Densities and distributions are important
2 Visualization methods fall off as variables get more
complex
3 Can we build on current methods to show conditional
dependencies and functional variables?

>>>>

\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{mindmap}

@@

\usepackage{array,multirow,graphicx}
\usepackage{subcaption}
\usepackage{subfiles}
\usepackage{url}
\usepackage{amsmath}

\title{Visualizing Conditional Dependencies}
\author{Hannah Aizenman}
\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
\submitted{}
\abstract{Many datasets are too large or complex to discern patterns simply through visualizing the observations in them, so researchers instead compute the distributions of select variables and explore under what conditions they occur. While there are many visualization techniques that preserve variable type, dimensionality, and multivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships. This survey presents a sampling of techniques for visualizing distributions and densities conditioned on variables that are categorical or quantitative, discrete or continuous, and potentially have or infer dimensional dependencies. 
}
\begin{document}
\makefrontmatter

\section{Introduction}
\label{sec:introduction}
Many datasets are multivariate and often contain a a mix of quantative and categorical variables. The Iris dataset \cite{fisher_use_1936-1, _uci_????} is a well known example of the type of data commonly found in machine learning literature. It is so wildely used as a sample dataset because it has 150 observations that have both quantative and categorical attributes. The version visualized in the introduction to this paper is obtained from the scikit-learn Python machine learning library \cite{scikit-learn}. The Iris dataset contains the following attributes:\\
\begin{tabular} {lllll}
Sepal length (cm) & Sepal width (cm) & Petal length (cm) & Petal width (cm) & Species name
\end{tabular}

\subsection{Formalization of Visualization Terms}

A first step in exploring the dataset is typically to make a visualization, but of what sort? Bertin posits that all visualizations at their core have two parts \cite{bertin_semiology_2011}:
\begin{description}
\item[invariant] - invariable common ground
\item[components] - a finite set of variational concepts
\end{description}

@@

\subsubsection{Bertin's Retinal Variables}
\begin{figure}
\includegraphics[width=1\textwidth]{intro/retinal_variables.png}
\caption{Bertin's retinal variables are a codification of how position, size, shape, color and texture can be used to illustrate variations in the components of a visualization. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} reproduced from Making Maps: A Visual Guide to Map Design for GIS \cite{krygier_making_2005}}
\label{fig:retinal_variables}
\end{figure}

@@

\subsubsection{Munzner's data semantics}

@@

\subsubsection{Conditional Probability}
\begin{figure}
\includegraphics[width=\textwidth]{intro/iris_parallel}
\caption{The parallel coordinates plot (PCP) of the Iris dataset shows the pairwise relationship between adjacent variables, for example sepal length and sepal width. The PCP plot can be used as proxy for illustrating conditional probability because it shows how the quantitative measurements distribute amongst the species. This is is seen both in the relative widths of each band of color and in the similarities in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands.}
\label{fig:iris_parallel}
\end{figure}

While the Iris dataset does not have any variables that seem to traditionally fit into the model of keys-as it has no time or space variables-an analogous look-up task for the Iris dataset is to filter on a component and then look at the probability distributions of the other variables in relation tp that key. In effect this treats all variables as potential keys into distributions of the other variables. In the parallel coordinates plot (PCP)\cite{claessen_flexible_2011,_nist/sematech_????,inselberg_plane_1985, wegman_hyperdimensional_1990} in figure~\ref{fig:iris_parallel}, each line is an observation in the dataset, and each color maps to a species of Iris. The color coding then allows for this visualization to be used to show how the four quantative variables (sepal length, sepal width, petal length, and petal width) distribute within each variable.

@@

wherein A is the observation vector $\left< \text{sepal length, sepal width, petal length, petal width} \right>$ and B in turn takes on each species of petal. The locations of the color and width of the colored band act as proxies for the distribution of each quantitative variable and the overall observation vector.

\subsubsection{Conditional Dependency}
\begin{figure}
\includegraphics[width=\textwidth]{intro/iris_observations.png}
\caption{The diagonal of the scatter matrix shows the distribution of each variable relative to the species type. This can be used to illustrate the conditional probability of each variable relative to a type of species. The scatter matrix plot shows every set of pairwise relationships in the dataset, color code by Iris species. Because the scatter matrix shows the co-occurance of two variables and is encoded with information about a third variable, it works well was a tool to explore a potential conditional dependency amongst the variables.}
\label{fig:iris_observations}
\end{figure}

As exemplified by the scatter matrix\cite{elmqvist_rolling_2008,l._wilkinson_high-dimensional_2006} in figure~\ref{fig:iris_observations}, there are a number of visualization techniques that can be used to explore conditional dependencies, but these methods are limited to datasets where the observations are discrete. There are visualization techniques for understanding the probability distributions of functional observations that could potentially be leveraged in conjunction with these methods to develop methods for visualizing conditional dependencies in a functional space.

\begin{figure}
\input{papermap.tikz}
\caption{The papers in this survey are organized by the type of observation and then by the type of measurements that makes up the observation.}
\label{fig:papermap}
\end{figure}

 This survey organizes the methods for exploring conditional dependency by observation type and data semantics, as shown in figure~\ref{fig:papermap}. In the functional space, the focus is on visualizing distributions because that is a necessary component of exploring conditional dependencies.

\section{Discrete Observations}
\subfile{sections/catset}
\subfile{sections/radclust}
\subfile{sections/pairwise}
\section{Continuous Observations}
\subfile{sections/quantKquantV}

@@

From climate models to Internet advertisement, a lot of modern datasets are very large and often structurally complex. And the questions being asked of these datasets are increasingly probabilistic-what is the distribution of observations, under what events do observations occur, and are there events that have to happen? The latter question is the one explored in conditional dependency visualizations. As explored in this paper, there are many techniques for exploring these dependencies in the discrete observation space, and there is a lot of work being done to understand distributions in the latter space. There is a space for future work that would leverage both to develop conditional dependency visualizations for functional datasets.

### 2018-01-31 paper/sections/quantKquantV.tex
newest version + repo clean up

\begin{document}
\section{Quantitative Variables Conditioned on Quantitative Densities}

\begin{figure}
  \includegraphics[width=\textwidth]{partial_dependence}
  \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots show how California housing prices are partially dependent on each x variable. The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
  \label{fig:partialdependence}
\end{figure}

One of the most straightforward ways to show dependencies is to plot one
variable against another, but those methods often exclude the effects of the
other variables in the datasets. Regression methods can be used to analyze the
ways in which one variable is partially dependent on many other variables in
the dataset\cite{_elements_2009}. As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.

While bar graphs and histograms can give a sense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham_40_2011}. While a simple box plot does not inherently show conditional dependencies, many derivations do.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot from
    Exploratory Data Analytics\cite{tukey_exploratory_1977} illustrates how the box plot shows the distribution of topology of states and
    the distribution of volcanoes. The strength of the box and whisker is that the
    reader can quickly compare the two distributions and learn, for example,
    that volcanoes tend to span a shorter range of average heights, but that the
    extremes are much further from the center. Essentially, the distributions act as references for the other displayed distributions.}
  \label{fig:boxplot}
\end{figure}

@@

While box plots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a timeseries
curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how close these functions are to each other--functional depth\cite{febrero_functional_2007},
bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
bivariate kernel density estimation\cite{scott_multivariate_1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{rob_j._hyndman_rainbow_2010}, but this method is unwieldy for larger
ensembles.

\begin{figure}
  \includegraphics[width=.75\linewidth]{contour_weather}
  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
  \label{fig:countour}
\end{figure}
Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution. A very common task is to compare ensembles, each a function of the same timeseries or geographic region. Contour box plots\cite{whitaker_contour_2013} aim to capture the variability of the
uncertainty via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members whiten all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand, which in this paper is temperature visualizations. The authors argue that contour box plots are an improvement over spaghetti plots \cite{luo_visualizing_2003} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns. Instead in
Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
the spaghetti, but removes the visual noise of the lines.

One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types. The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman_eyes_1996} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core. 
\end{document}

>>>>

\begin{document}
\subsection{Variations on Boxplots}

While bar graphs and histograms can give a sense of the variability in data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham_40_2011}.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot from
    The box plot shows the distribution of topology of states and
    the distribution of volcanoes. The strength of the box and whisker is that the
    reader can quickly compare the two distributions and learn, for example,
    that volcanoes tend to span a shorter range of average heights, but that the
    extremes are much further from the center. Essentially, the distributions act as references for the other displayed distributions. Figure is from Exploratory Data Analytics\cite{tukey_exploratory_1977} }
  \label{fig:boxplot}
\end{figure}

@@

\subsubsection{Boxplots for Curves}
While box plots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a functional densities, drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
One such example is a timeseries curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how close these functions are to each other--depth\cite{febrero_functional_2007,lopez-pintado_functional_2007},
bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
bivariate kernel density estimation\cite{scott_multivariate_1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{rob_j._hyndman_rainbow_2010}, but this method is unwieldy for larger
ensembles.

\begin{figure}
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.8]{spaghetti}
    \caption{Spaghetti plot of ensemble of 500mb and -15C temperature isocountours.}
    \label{fig:spaghetti}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.5]{contour_weather}
    \caption{This countour boxplot is showing the distribution of a temperature field at 500mb and -15C. The mean field is in yellow, the median in purple, red as outlier cut offs, and envelops for the minimum and maxmimum bands.}
    \label{fig:contour}
    \end{subfigure}
    \caption{Both figures are from Contour boxplots: A method for characterizing uncertainty in feature sets from simulation ensembles\cite{whitaker_contour_2013}}
\end{figure}

Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited in the variability they show. A very common task is to compare ensembles, each a function of the same timeseries or geographic region. Contour box plots\cite{whitaker_contour_2013} aim to capture the variability of the
uncertainty via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members whiten all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand, which in this paper is temperature visualizations. As illustrated by figure~\ref{fig:spaghetti}, contour boxplots are an improvement over spaghetti plots \cite{luo_visualizing_2003, whitaker_contour_2013} because a lot of the nosiness is filtered out.  In figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of the spaghetti plot in figure~\ref{fig:spaghetti}, but removes the visual noise of the lines.

\subsubsection{Boxplots for Surfaces}
\begin{figure}
  \includegraphics[width=1\linewidth]{surfacebox.png}
  \caption{The large colored surface in the middle is the median surfac. The surfaces on top are minimum and maximum envelopes. The graphs at the bottom visualize the distribution of band-depths of surfaces. The 3 surfaces on the right have been identified as outliers by the tool. This figure is from Surface boxplots \cite{genton_surface_2014}}
  \label{fig:surface}
\end{figure}
Another application of this band depth concept is to surfaces such as collections of images or functional data generated by models. In figure~\ref{fig:surface}, Genton et al. built a surface boxplot visualization tool with multiple panes\cite{genton_surface_2014}. Like a traditional boxplot, they display the median surface from the dataset, minimum and maximum surfaces, and they have a 3D view that gives a sense of how surfaces are distributed. The tool also provides histograms of the band depth to give a 2d view on how the surfaces are distributed.

\end{document}
